Report for Qualifying Exam

### 2024-08-08 src/sections/introduction.tex
minor changes

\section{Introduction} \label{sec:introduction}
In the digital age, online information seekers frequently encounter \emph{search results} laden with \emph{misinformation}, \emph{biased narratives}, and \emph{conspiracy theories}.
This exposure can potentially lead users to accept and propagate false information.
However, this phenomenon is not universal across all search queries.
Certain searches, known as \textit{data voids} are particularly susceptible to these issues.
\\
\textit{Data voids} occur when searching for terms that yield limited, non-existent, or highly problematic relevant information.
Unlike common searches that produce abundant data, queries falling into data voids may return no results or present irrelevant and often inaccurate information.
Data voids can be mainly categorized into two main types:
\begin{itemize}
    \item newly coined terms with no established information base, and
    \item existing terms whose meanings have evolved over time.
\end{itemize}
This report focuses on the latter, examining data voids where context drift occurs.
The following papers has been selected to explore this phenomenon.
\begin{itemize}
    \item Statistically Significant Detection of Linguistic Change.
    ACM WWW 2014 ~\cite{kulkarni2014statisticallysignificantdetectionlinguistic}
    \item Diachronic Word Embeddings Reveal Statistical Laws of Semantic Change.
    ACL 2016 ~\cite{hamilton-etal-2016-diachronic}
\end{itemize}

@@

\subsection{Example of a data void?(Migrant Caravan)}
Let's understand data void through an example of a term - \emph{migrant caravan}
The term ‘migrant caravan’ originally described a group of migrants traveling together for safety and mutual support.
However, its usage, particularly in political discussions, has taken on a more negative tone.
Today, it is commonly associated with large groups of migrants, often from Central America, journeying through Mexico to reach the United States.

\subsection{Lifecycle of data void (from long tail to spike)}
\subsection{Types of Data Void}
\subsection{Technical Challenge}
analyzing data void involves context drift over historical time. (How context drift happens with data void.)

>>>>

\section{Introduction} \label{sec:introduction}
Language is a dynamic, living entity that continually adapts to reflect the changing world around us.
With the evolving language, words take on new meanings or connotations with technological advancements,
social and cultural shifts, generational differences, globalization, political and social movements, etc.
The particular words when get emerged in political context are known as data voids.
Since, they have the behavior of changing meaning over time.
There can be "voids" in the meaning of words or concepts, especially in specific contexts like politics.
When a word's meaning shifts rapidly or becomes ambiguous, it can create a kind of "data void" for analysis or understanding.

Data voids in political contexts often occur when terms or phrases take on new or altered meanings,
particularly when these changes are driven by political actors or movements.

This type of data void is especially significant because:

Rapid meaning shifts: Political events or campaigns can cause sudden changes in how certain terms are used or interpreted.
Intentional manipulation: Political actors may deliberately redefine terms or introduce new phrases to frame debates in their favor.
Polarization of language: Words or phrases can take on drastically different meanings in different political contexts or for different groups.
Dog whistles: Some terms may develop coded political meanings that aren't immediately apparent to all audiences.
Weaponization of ambiguity: The lack of clear, authoritative information about a term's new political meaning can be exploited to spread misinformation or propaganda.

Media manipulators often manipulate language to shape public opinion.
Words are carefully chosen to present a particular viewpoint.
Changes in political ideologies can alter the meaning of terms.
Broader societal shifts influence political discourse.

These elements can rapidly change the connotations and denotations of words, making them challenging to track and analyze.

data void isn't just about a lack of information, but about the potential for confusion, manipulation, or misinformation as terms are redefined or recontextualized in the political sphere.

This type of data void highlights the dynamic nature of language in political discourse and the challenges it presents for maintaining shared understanding in public debates.
It also underscores the importance of being aware of how language can be used as a political tool.

The following papers has been selected to explore this phenomenon.
\begin{itemize}
    \item Statistically Significant Detection of Linguistic Change.
    ACM WWW 2014 ~\cite{kulkarni2014statisticallysignificantdetectionlinguistic}
    \item Diachronic Word Embeddings Reveal Statistical Laws of Semantic Change.
    ACL 2016 ~\cite{hamilton-etal-2016-diachronic}
\end{itemize}

@@

### 2024-08-08 src/sections/paper_hamilton.tex
minor changes

@@

\subsection{Constructing Word Embeddings.}\label{subsec:constructing-word-embeddings.}
The authors employ three methods to construct word embeddings for different time periods.
Initially, they create embeddings for each distinct period and then align these embeddings over time to maintain consistency.
To quantify semantic change, they use various metrics.
Specifically, they utilize Singular Value Decomposition (SVD), Positive Pointwise Mutual Information (PPMI), and Skip-Gram with Negative Sampling (SGNS).
These distributional techniques represent each word by a vector that encapsulates information about the word’s co-occurrence statistics.

\begin{enumerate}
    \item
\para{PPMI: }
PPMI is a statistical measure used to quantify the association between a word and its context within a corpus.
In the paper, it is used to create word embeddings by capturing the strength of association between words and their co-occurring contexts over time.

\begin{itemize}
    \item \para{Co-occurrence Matrix Construction.}
    First, a co-occurrence matrix is built, where each entry represents the frequency with which a word (target) and its context (typically a window of neighboring words) appear together in the corpus.
    \item \para{Joint Probability Calculation.}
    The joint probability of a target word $w_i$ and a context word $c_j$ is calculated by dividing the co-occurrence frequency of $w_i$ and $c_j$ by the total number of word pairs in the corpus.
    \item \para{Marginal Probability Calculation.}
    The marginal probabilities of the target word $w_i$ and the context word $c_j$ are calculated based on the total occurrences of each word in the corpus.
    \item \para{Calculating PPMI.}
    The Positive Pointwise Mutual Information (PPMI) between the target word $w_i$ and the context word $c_j$ is calculated using the formula:

    \begin{equation}
    \mathbf{M}^{\text{PPMI}}_{i,j} = \max \left\{ \log \left( \frac{\hat{p}(w_i, c_j)}{\hat{p}(w_i) \hat{p}(c_j)} \right) - \alpha, 0 \right\},
    \label{eq:ppmi}
    \end{equation}

    where:
    \begin{itemize}
    \item $\hat{p}(w_i, c_j)$ is the estimated \emph{joint probability} of the target word $w_i$ and the context word $c_j$ occurring together.
    \item $\hat{p}(w_i)$ and $\hat{p}(c_j)$ are the estimated \emph{marginal probabilities} of the target word $w_i$ and the context word $c_j$, respectively.
    \item $\alpha > 0$ is a positive constant used to smooth the values and prevent extreme positive values in the PPMI calculation.
    \end{itemize}
    This formula refines the PPMI measure by incorporating $\alpha$ to adjust the log probability ratio.
    The inclusion of $\alpha$ helps manage the sparseness of the data and the impact of low-frequency events, making the measure more robust and stable.
\end{itemize}

    \item
\para{SVD: }
SVD is a mathematical technique that factorizes a matrix into three other matrices, revealing important features of the original matrix.
In the context of word embeddings, SVD helps to reduce the dimensionality of the data while preserving the essential structure and patterns in word co-occurrences.
SVD decomposes the co-occurrence matrix into three matrices in the following manner:

\begin{equation}
\mathbf{w}_i^{\text{SVD}} = (\mathbf{U} \Sigma^\gamma)_i,
\label{eq:svd}
\end{equation}

@@

    \item
\para{SGNS: }
SGNS is a technique used to learn high-quality word vectors by training on large corpora.
The primary goal of SGNS is to learn word embeddings such that words that appear in similar contexts have similar vector representations.
\\
For a given target word  w  and a context word  c , the model tries to maximize the probability  P(c|w)

\begin{equation}
\hat{p}(c_i \mid w_i) \propto \exp(\mathbf{w}_i^{\text{SGNS}} \cdot \mathbf{c}_j^{\text{SGNS}}),
\label{eq:sgns}
\end{equation}

where:

\begin{itemize}
    \item $\hat{p}(c_i \mid w_i)$ is the estimated probability of a context word $c_i$ given a target word $w_i$.
    \item $\mathbf{w}_i^{\text{SGNS}}$ is the vector representation of the target word $w_i$ in the SGNS embedding space.
    \item $\mathbf{c}_j^{\text{SGNS}}$ is the vector representation of the context word $c_j$ in the SGNS embedding space.
\end{itemize}

\\
The probability $\hat{p}(c_i \mid w_i)$ that a word $c_i$ is a context word of $w_i$ is proportional to the exponential of the dot product of their corresponding vector representations.
The dot product $\mathbf{w}_i^{\text{SGNS}} \cdot \mathbf{c}_j^{\text{SGNS}}$ measures the similarity between the target word and the context word in the vector space.
A higher dot product indicates that the words are more likely to co-occur, leading to a higher estimated probability.\\

The model uses this probability estimation to distinguish true word-context pairs from randomly sampled "negative" pairs.
The training objective is to maximize the probability of observed (target, context) pairs and minimize the probability of randomly sampled pairs, thereby learning embeddings that reflect the semantic relationships between words.

\end{enumerate}

\subsection{Aligning Embeddings}\label{subsec:aligning-embeddings}
Since the embeddings are trained independently for each time period, their vector spaces may not be directly comparable.
The paper addresses this by aligning the embeddings from different time periods.
They aligned the emebddings across time periods using orthogonal Procrustes, which transforms the embeddings to minimize the distances between corresponding words in different periods.
This step ensures that similar meanings are represented similarly across time.

\subsection{Measuring Semantic Change}\label{subsec:measuring-semantic-change}
Once the embeddings are aligned, following methods are performed to quantify how much the meaning of word changes over time.
\begin{itemize}
    \item \para{Pair-wise similarity time-series}
        This method measures changes in the similarity between pairs of words over different time periods using cosine similarity.
        \begin{equation}
            s^{(t)}(w_i, w_j) = \text{cos-sim}(w_i^{(t)}, w_j^{(t)})
            \label{eq:equation2}
        \end{equation}
        The Spearman correlation $(\rho)$ is employed to measure the relationship between the similarity scores of word pairs and time.
        This non-parametric method assesses whether the similarity series shows a significant increase or decrease over time, which is crucial for understanding semantic shifts.

    \item \para{Measuring semantic displacement}
        This method quantifies how much a word's meaning has changed by calculating the displacement of its vector representation across different time points.
        They use the aligned word vectors to compute the \emph{semantic displacement} that a word has undergone during a certain time-period.
        It is measured by calculating the distance between the vector representations of a word in different time periods.
        For a given word $w$, if $\mathbf{w}^{(t1)}$ and $\mathbf{w}^{(t2)}$ are its embeddings in two time periods $t1$ and $t2$, respectively, then the semantic displacement $\Delta$ is computed as:
        \begin{equation}
            \Delta(w) = \text{cos-dist}(\mathbf{w}^{(t1)}, \mathbf{w}^{(t2)}).\label{eq:equation3}
        \end{equation}
\end{itemize}

\subsection{Evaluation of various approaches}
They compared the different approaches discussed in \Cref{subsec:methods-for-quantifying-semantic-change} on \emph{synchronic accuracy} (similarity within time-period) and \emph{diachronic validity} (semantic changes over time).
\begin{itemize}
    \item \para{Synchronic Accuracy}
    Synchronic accuracy refers to the accuracy with which word embeddings capture the semantic relationships and meanings of words at a \emph{specific point in time}.
    It assesses how well the models represent the semantic relationships among words in a given time period.
    SVD performed best here.
    \item \para{Diachronic Validity}
    Diachronic validity evaluates how effectively the methods can detect and quantify changes in \emph{meaning over time}.
    It involves testing the methods against historical data to see if they can accurately identify known shifts in word meanings.
    SGNS perfomed best on both of the tasks:
    \begin{itemize}
        \item \textbf{Detecting Known Shifts.}
        The goal in this task is for the methods to correctly capture whether pairs of words moved closer or further apart in semantic space during a pre-determined time-period.

        \item \textbf{Discovering Shifts from data.}
        Tested whether the methods discover reasonable shifts by examining the top-10 words that changed the most from the 1900s to the 1990s according to the semantic displacement metric (\Cref{subsec:methods-for-quantifying-semantic-change}).
    \end{itemize}

@@

\subsection{Takeaways}
This paper explores how different types of word embedding models, such as PPMI, SVD, and SGNS, can be used to study diachronic shifts in word meanings.
It demonstrates the use of metrics like Spearman correlation to quantify semantic drift and employs Procrustes alignment techniques to align word embeddings across different time periods.
By analyzing these aligned embeddings, the authors identify two key statistical laws of semantic change.
These findings provide a quantitative framework for understanding how and why languages evolve over time.

>>>>

In our work on data voids, we have recognized that word frequency significantly impacts their identification.
For instance, the term `sandy' saw a spike in usage during the Hurricane Sandy event.
Similarly, terms associated with political agendas also experience increased usage during relevant events.
This paper provides insights into how frequency influences the evolution of word meanings.

Additionally, our study identifies data voids involving polysemous terms, such as `migrant caravan.'
This term began to change in meaning when FOX News used it to describe people moving from the Mexico border towards the US, and it continued to evolve with new groups of migrants.
Understanding these kinds of data voids is crucial, and this paper helps us identify subtle semantic changes in words with multiple meanings,
enhancing our understanding of how these changes relate to frequency and context.

@@

\subsection{Constructing Word Embeddings.}\label{subsec:constructing-word-embeddings}
The authors employ three methods to construct word embeddings for different time periods.
Initially, they create embeddings for each distinct period and then align these embeddings over time to maintain consistency.
To quantify semantic change, they use various metrics.
Specifically, they utilize Singular Value Decomposition (SVD), Positive Pointwise Mutual Information (PPMI), and Skip-Gram with Negative Sampling (SGNS).
These distributional techniques represent each word by a vector that encapsulates information about the word’s co-occurrence statistics.

\para{Positive Pointwise Mutual Information (PPMI). }
PPMI is a statistical measure used to quantify the association between a word and its context within a corpus.
In the paper, it is used to create word embeddings by capturing the strength of association between words and their co-occurring contexts over time.

First, a co-occurrence matrix is built, where each entry represents the frequency with which a word (target) and its context (typically a window of neighboring words) appear together in the corpus.
The \emph{joint probability} of a target word $w_i$ and a context word $c_j$ is calculated by dividing the co-occurrence frequency of $w_i$ and $c_j$ by the total number of word pairs in the corpus.
The \emph{marginal probability} of the target word $w_i$ and the context word $c_j$ are calculated based on the total occurrences of each word in the corpus.
The Positive Pointwise Mutual Information (PPMI) between the target word $w_i$ and the context word $c_j$ is calculated using the formula:
\begin{equation}
\mathbf{M}^{\text{PPMI}}_{i,j} = \max \left\{ \log \left( \frac{\hat{p}(w_i, c_j)}{\hat{p}(w_i) \hat{p}(c_j)} \right) - \alpha, 0 \right\},
\label{eq:ppmi}
\end{equation}

where:
\begin{itemize}
\item $\hat{p}(w_i, c_j)$ is the \emph{joint probability} of the target word $w_i$ and the context word $c_j$ occurring together.
\item $\hat{p}(w_i)$ and $\hat{p}(c_j)$ are the \emph{marginal probabilities} of the target word $w_i$ and the context word $c_j$, respectively.
\item $\alpha > 0$ is a positive constant used to smooth the values and prevent extreme positive values in the PPMI calculation.
\end{itemize}

This formula refines the PPMI measure by incorporating $\alpha$ to adjust the log probability ratio.
The inclusion of $\alpha$ helps manage the sparseness of the data and the impact of low-frequency events, making the measure more robust and stable.

\para{Singular Value Decomposition (SVD). }
SVD is a mathematical technique that factorizes a matrix into three other matrices, revealing important features of the original matrix.
In the context of word embeddings, SVD helps to reduce the dimensionality of the data while preserving the essential structure and patterns in word co-occurrences.
SVD decomposes the co-occurrence matrix into three matrices in the following manner:
\begin{equation}
\mathbf{w}_i^{\text{SVD}} = (\mathbf{U} \Sigma^\gamma)_i,
\label{eq:svd}
\end{equation}

@@

\para{Skip-Gram with Negative Sampling (SGNS). }
SGNS is a technique used to learn high-quality word vectors by training on large corpora.
The primary goal of SGNS is to learn word embeddings such that words that appear in similar contexts have similar vector representations.
For a given target word $w$ and a context word $c$, the model tries to maximize the probability $P(c|w)$
\begin{equation}
\hat{p}(c_i \mid w_i) \propto \exp(\mathbf{w}_i^{\text{SGNS}} \cdot \mathbf{c}_j^{\text{SGNS}}),
\label{eq:sgns}
\end{equation}

where:
\begin{itemize}
    \item $\hat{p}(c_i \mid w_i)$ is the estimated probability of a context word $c_i$ given a target word $w_i$.
    \item $\mathbf{w}_i^{\text{SGNS}}$ is the vector representation of the target word $w_i$ in the SGNS embedding space.
    \item $\mathbf{c}_j^{\text{SGNS}}$ is the vector representation of the context word $c_j$ in the SGNS embedding space.
\end{itemize}

The probability $\hat{p}(c_i \mid w_i)$ that a word $c_i$ is a context word of $w_i$ is proportional to the exponential of the dot product of their corresponding vector representations.
The dot product $\mathbf{w}_i^{\text{SGNS}} \cdot \mathbf{c}_j^{\text{SGNS}}$ measures the similarity between the target word and the context word in the vector space.
A higher dot product indicates that the words are more likely to co-occur, leading to a higher estimated probability.

The model uses this probability estimation to distinguish true word-context pairs from randomly sampled negative pairs.
The training objective is to maximize the probability of observed (target, context) pairs and minimize the probability of randomly sampled pairs, thereby learning embeddings that reflect the semantic relationships between words.

\para{Aligning Embeddings.}
Since the embeddings are trained independently for each time period, their vector spaces may not be directly comparable.
The paper addresses this by aligning the embeddings from different time periods.
They aligned the emebddings across time periods, which transforms the embeddings to minimize the distances between corresponding words in different periods.
This step ensures that similar meanings are represented similarly across time.

\subsection{Takeaways}\label{subsec:hamilton_takeaways1}
The authors created three types of word embeddings using PPMI, SVD, and SGNS, each capturing different information.
PPMI measures how closely related two words are, resulting in sparse, high-dimensional vectors.
SVD simplifies this information by reducing the dimensionality of the PPMI matrix into dense, low-dimensional vectors.
Lastly, SGNS learns embeddings by predicting surrounding words in a given context, uncovering deeper semantic and contextual nuances that the other methods may not capture.
Additionally, they employed alignment technique to align word embeddings across different time periods, ensuring consistency and enabling the analysis of semantic shifts over time.

%\subsection{Measuring Semantic Change}\label{subsec:measuring-semantic-change}
%Once the embeddings are aligned, following methods are performed to quantify how much the meaning of word changes over time.
%\begin{itemize}
%    \item \para{Pair-wise similarity time-series}
%        This method measures changes in the similarity between pairs of words over different time periods using cosine similarity.
%        \begin{equation}
%            s^{(t)}(w_i, w_j) = \text{cos-sim}(w_i^{(t)}, w_j^{(t)})
%            \label{eq:equation2}
%        \end{equation}
%        The Spearman correlation $(\rho)$ is employed to measure the relationship between the similarity scores of word pairs and time.
%        This non-parametric method assesses whether the similarity series shows a significant increase or decrease over time, which is crucial for understanding semantic shifts.
%
%    \item \para{Measuring semantic displacement}
%        This method quantifies how much a word's meaning has changed by calculating the displacement of its vector representation across different time points.
%        They use the aligned word vectors to compute the \emph{semantic displacement} that a word has undergone during a certain time-period.
%        It is measured by calculating the distance between the vector representations of a word in different time periods.
%        For a given word $w$, if $\mathbf{w}^{(t1)}$ and $\mathbf{w}^{(t2)}$ are its embeddings in two time periods $t1$ and $t2$, respectively, then the semantic displacement $\Delta$ is computed as:
%        \begin{equation}
%            \Delta(w) = \text{cos-dist}(\mathbf{w}^{(t1)}, \mathbf{w}^{(t2)}).\label{eq:equation3}
%        \end{equation}
%\end{itemize}

\subsection{Evaluation \& Results}
They compared the different approaches discussed in \Cref{subsec:constructing-word-embeddings} on \emph{synchronic accuracy} (similarity within time-period) and \emph{diachronic validity} (semantic changes over time).
\begin{itemize}
    \item \para{Synchronic Accuracy}
    Synchronic accuracy refers to the accuracy with which word embeddings capture the semantic relationships and meanings of words at a \emph{specific point in time}.
    It assesses how well the models represent the semantic relationships among words in a given time period.
    SVD performed best here.
    \item \para{Diachronic Validity}
    Diachronic validity evaluates how effectively the methods can detect and quantify changes in \emph{meaning over time}.
    It involves testing the methods against historical data to see if they can accurately identify known shifts in word meanings.
    The results indicated that SGNS was particularly effective in capturing these shifts, followed by SVD and PPMI.
    \begin{itemize}
        \item \textbf{Detecting Known Shifts.}
        The goal in this task is for the methods to correctly capture whether pairs of words moved closer or further apart in semantic space during a pre-determined time-period.

        \item \textbf{Discovering Shifts from data.}
        Tested whether the methods discover reasonable shifts by examining the top-10 words that changed the most from the 1900s to the 1990s according to the semantic displacement metric.
    \end{itemize}

@@

\subsection{Takeaways}
This paper explores how different types of word embedding models, such as PPMI, SVD, and SGNS, can be used to study diachronic shifts in word meanings.
By analyzing these aligned embeddings, the authors identify two key statistical laws of semantic change.
These findings provide a quantitative framework for understanding how and why languages evolve over time.

### 2024-08-08 src/sections/paper_kulkarni.tex
minor changes

\begin{wrapfigure}{r}{0.3\textwidth}
    \centering
    \vspace{-1em}
    \includegraphics[width=0.35\textwidth]{figures/frequency-gay}
    \vspace*{-0.7cm}
    \caption{Changes in frequency of 'gay' (taken from \cite{kulkarni2014statisticallysignificantdetectionlinguistic}).}
    \label{fig:example-gay}
\end{wrapfigure}

@@

\\
Tools such as the Google Books Ngram Viewer and Google Trends are used for this purpose, as they provide large datasets and visual representations of word usage across different timeframes.
They calculate the change in probability of a word appearing over time.
\begin{equation}
\mathbf{T_t(w)} = \log \frac{#(w \in C_t)}{|C_t|}
\label{eq:equation}
\end{equation}

@@

\subsection{Takeaways}\label{subsec:takeaways}
For the first research question, Kulkarni et al.\ (2015) explored multiple dimensions of word usage change over time.
To address this, they constructed three types of time series: frequency-based, syntactic, and distributional.
Among these, the distributional time series proved most insightful, as it captures the contextual nuances of word use.
This approach utilized word embeddings to capture the semantic context of a word within specific time periods.
Subsequently, these embeddings were aligned into a unified space,
and it relied on assumptions about word stability and embedding space linearity.
These limitations affect the method's ability to fully capture complex linguistic changes.
Despite these challenges, the distributional approach provides a foundational understanding of language dynamics.

>>>>

\begin{wrapfigure}{r}{0.3\textwidth}
    \centering
    \vspace{-3em}
    \includegraphics[width=0.35\textwidth]{figures/frequency-gay}
    \vspace*{-1cm}
    \caption{Changes in frequency of 'gay' (taken from \cite{kulkarni2014statisticallysignificantdetectionlinguistic}).}
    \label{fig:example-gay}
\end{wrapfigure}

@@

Tools such as the Google Books Ngram Viewer and Google Trends are used for this purpose, as they provide large datasets and visual representations of word usage across different timeframes.
They calculate the change in probability of a word appearing over time.
\begin{equation}
\mathbf{T_t(w)} = \log \frac{#(w \in C_t)}{|C_t|}
\label{eq:equation}
\end{equation}

@@

\subsection{Takeaways}\label{subsec:takeaways}
For the first research question, Kulkarni et al.\ (2015) explored multiple dimensions of word usage change over time.
To address this, they constructed three types of time series: frequency-based, syntactic, and distributional.
Among these, the distributional time series proved most insightful, as it captures the contextual usage of word.
This approach utilized word embeddings to capture the semantic context of a word within specific time periods.
Subsequently, these embeddings were aligned into a unified space,
and it relied on assumptions about word stability and embedding space linearity.
